\section{Trying around}

\subsection{VI/BVI}

Memory: 2 numbers for every state

\begin{itemize}
	\item Initialize.
	
	L = 1 if target else 0
	
	U = 0 if can't reach else 1
	\item Repeat until U-L$<\varepsilon$ for initial state
	\begin{itemize}
		\item Bellman update for U and L on whole state space.
	\end{itemize}
\end{itemize}

\subsection{BRTDP}

Memory: 2 numbers for every seen state, potentially for every state.

\nextstate: Det, Unif, Prob, MaxDiff.

\stuck: PathLen (explored states, all states, iters, log, exp...), seenTwice, FSMC.
May have false positives, but no false negatives.

\begin{itemize}
	\item Repeat until $U-L<\varepsilon$ for initial state
	\begin{itemize}
		\item Repeat until last state in path is target or \stuck.
		\begin{itemize}
		\item Simulate: Repeatedly pick successor according to \nextstate. Save path (list of states).
		\end{itemize}
	\item Bellman update for U and L on states in path. 
	
	New states always have L=0 and U=1, except target states have L=1 and we need some way to detect U=0. Using FSMC, we can set BSCC to 0. We can also check if we are in a BSCC when STUCK.
	
	\item Check BSCC without building full partial model (look for outgoing edges between the looping states)
	\end{itemize}
\end{itemize}

\textcolor{red}{Compare to other STUCK heuristics (FSMC p2, para 2). Compare to importance sampling/splitting - see Modest papers.}

\subsection{SMC}

Memory: Number for initial state, path to detect candidate.

\nextstate: Prob.

\stuck: FSMC.
Needs to be exact

\begin{itemize}
	\item Repeat enough times (enough depends on n and p and stuff)
	\begin{itemize}
		\item Repeat until last state in path is target or \stuck.
		\begin{itemize}
			\item Simulate: Repeatedly pick successor according to \nextstate. Save path (list of states).
		\end{itemize}
		\item Remember 0/1 depending on target or \stuck.
	\end{itemize}
\end{itemize}

Result is sum of remembered 0/1 divided by how often we simulated.


\subsection{Precise SMC}

Use other SMC to guide BRTDP? We already have something better than 1.3 by using MaxDiff.

\subsection{Experiments}

MODEST/PRISM-SMC/PRISM (Standard)/Our BRTDP/PLASMA LAB/STORM